---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tax4sa

<!-- badges: start -->
<!-- badges: end -->

This is a minimal package to help with the compilation and analysis of tax data in South Africa. The package only contains three main sets of data, three functions and the personal income tax tables from 1995/96 to 2021/22. 

The data includes monthly tax revenue collections of the South African Revenue Service (SARS) as published by the [National Treasury of South Africa](http://www.treasury.gov.za/comm_media/press/monthly/default.aspx) in a dataframe named `SARS`, data from the Quarterly Bulletin published by the [South African Reserve Bank](https://www.resbank.co.za/en/home/publications/quarterly-bulletin1/download-information-from-xlsx-data-files) in a dataframe named `SARB` and regularly updated statistics from [Statistics South Africa](http://www.statssa.gov.za/?page_id=1847) in a dataframe named `STATSSA`.

The three functions and the personal income tax tables are intended to help with calculating tax liabilities, particularly when used with the [administrative data from SARS](https://sa-tied.wider.unu.edu/sites/default/files/pdf/SATIED_WP36_Ebrahim_Axelson_March_2019.pdf).

## Installation

You can install the package from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("chrisaxelson/tax4sa")
```
## Example

The data can be accessed by directly entering either `SARS` or `STATSSA` and is in a tidy format to ease analysis within R. `SARB_descriptions` and `STATSSA_descriptions` are also available to help with the details of each variable in those two sets of data.  

```{r example, message = FALSE}
library(tax4sa)
library(dplyr)
library(knitr)

# Check revenue data
SARS %>% 
  filter(Tax == "Total tax revenue (gross)") %>% 
  tail(5) %>% 
  kable()

# Look for SARB economic data on GDP
SARB_descriptions %>% 
  filter(grepl("Gross domestic product at market prices", Description), Frequency == "K1") %>%
  kable()

SARB %>% 
  filter(Code == "KBP6006K") %>% 
  tail(5) %>% 
  kable()

# Look for STATSSA inflation data
STATSSA_descriptions %>% 
  filter(grepl("Consumer Price Index", H02), H04 == "All Items") %>%
  select(H01, H02, H03, H04, H13) %>% 
  kable()

STATSSA %>% 
  filter(Code == "CPT00000") %>% 
  tail(5) %>% 
  kable()


```

The aim is to update the data monthly and the data structure should hopefully stay the same to allow for automated updates. 

```{r revenue, message = FALSE}
library(dplyr)
library(tsibble)
library(ggplot2)
library(scales)

Total_revenue <- SARS %>% 
  filter(Tax == "Total tax revenue (gross)") %>% 
  mutate(Year_month = yearmonth(paste(Year, Month)))

ggplot(Total_revenue, aes(x = Year_month, y = Revenue)) +
  geom_line(color = "darkblue") + 
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(axis.title.x = element_blank())
  
```

The three functions are `tax_calculation`, `pit` and `pit_manual`. The first is a generic function to apply a tax table to a value, while the latter two specifically calculate the personal income tax liability in South Africa. `pit_manual` allows for a custom tax table to be applied to cater for modelling the impacts of changes in the personal income tax tables. The package includes a list of historical tax tables to be used in the calculations. 

```{r tax, message = FALSE}
# Accessing tax tables
tax_calculation(100000, Tax_tables$PIT_brackets_2021)

# Calculate personal income tax
pit(income = 1000000, age = 53, mtc = 2550, tax_year = 2021)

# Same calculation in a relatively large dataframe with differing variables
individuals <- 1e6
df <- data.frame(Taxable_income = round(runif(individuals, 0, 3000000),0),
                 Age = round(runif(individuals, 18, 80),0),
                 MTC = round(runif(individuals, 0, 6000), 0),
                 Tax_year = round(runif(individuals, 2014, 2020), 0))

system.time({
  df <- df %>% 
    mutate(Simulated_tax = pit(Taxable_income, Age, MTC, Tax_year))
})
```
