---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tax4sa

<!-- badges: start -->
<!-- badges: end -->

This is a minimal package to help with the compilation and analysis of tax data in South Africa. The package only contains two main sets of data and three functions. 

The data includes monthly tax revenue collections of the South African Revenue Service (SARS) as published by the [National Treasury of South Africa](http://www.treasury.gov.za/comm_media/press/monthly/default.aspx) in a dataframe named `SARS` and data from the Quarterly Bulletin published by the [South African Reserve Bank](https://www.resbank.co.za/en/home/publications/quarterly-bulletin1/download-information-from-xlsx-data-files).

The functions are intended to help with calculating tax liabilities, particularly when used with the [administrative data from SARS](https://sa-tied.wider.unu.edu/sites/default/files/pdf/SATIED_WP36_Ebrahim_Axelson_March_2019.pdf).

## Installation

You can install the package from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("chrisaxelson/tax4sa")
```
## Example

The data can be accessed by directly entering either `SARS` or `SARB` and is in a tidy format to ease analysis within R. `SARB_descriptions` is also available to help with the Quarterly Bulletin codes,  

```{r example, message = FALSE}
library(tax4sa)
library(knitr)

# Check data
kable(SARS[1:5,])
kable(SARB[1:5,])
kable(SARB_descriptions[1:3,])
```

The SARS data tries to use the underlying data names as far as possible, and the names of the same revenue items are adjusted to be consistent across previous years. For example, total tax revenue is stated as "Total tax revenue (gross)".

```{r revenue, message = FALSE}
library(dplyr)
library(tsibble)
library(ggplot2)
library(scales)

Total_revenue <- SARS %>% 
  filter(Tax == "Total tax revenue (gross)") %>% 
  mutate(Year_month = yearmonth(paste(Year, Month)))


ggplot(Total_revenue, aes(x = Year_month, y = Revenue)) +
  geom_line(color = "darkblue") + 
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(axis.title.x = element_blank())
  
```

The three functions are `tax_calculation`, `pit` and `pit_manual`. The first is a generic function to apply a tax table to a value, while the latter two specifically calculate the personal income tax liability in South Africa. `pit_manual` allows for a custom tax table to be applied to cater for modelling the impacts of changes in the personal income tax tables. The package includes a list of historical tax tables to be used in the calculations. 

```{r tax, message = FALSE}
# Accessing tax tables
tax_calculation(100000, Tax_tables$PIT_brackets_2021)

# Calculate personal income tax
pit(income = 1000000, age = 53, mtc = 2550, taxyear = 2021)

# Same calculation in a relatively large dataframe with differing variables
individuals <- 1e6
df <- data.frame(Taxable_income = round(runif(individuals, 0, 3000000),0),
                 Age = round(runif(individuals, 18, 80),0),
                 MTC = round(runif(individuals, 0, 6000), 0),
                 Tax_year = round(runif(individuals, 2014, 2020), 0))

system.time({
  df$Simulated_tax <- pit(df$Taxable_income, df$Age, df$MTC, df$Tax_year)
})
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub!
